Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
edit in progress (memoization)

%\subsection{MRFy Implementation}

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality.
They form just one component of our new tool, MRFy.

A hidden Markov model uses local information about
amino-acid sequences.
But real proteins are three-dimensional,
and when a real protein folds, groups of amino acids
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
We~are studying adjacent groups called \emph{beta strands}, which
can become hydrogen-bonded to each other,
and are effectively ``stuck together.''

A~mutation in a paired beta strand may break the pairing, in which
 case the mutated protein will not fold properly and is unlikely to
 survive natural selection.
But a mutation in a paired beta strand \emph{can} survive if it is
 accompanied by a compensatory mutation in the strand with which it is
 paired. 
The probability of finding amino acid $x_i$ in column~$j$ of an alignment may
 therefore depend on the amino acid in the paired column
 $\pairedwith j$.
This nonlocal interaction changes
Viterbi's equations:
each term $V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
The distance between columns~$j$ and $\pairedwith j$ can be as small
 as a few columns or as large as a few hundreds of columns.

The approach currently taken in computational biology
is~to model each pair of beta strands not with two hidden Markov
models, 
but with a pair of sequences of match states.
Each of the subsequences \emph{between} paired beta strands is then
modeled with an ordinary hidden Markov model.
The~combined model, as shown in \figref{mrf}, is called a
\textit{Markov random field}. 
Because of nonlocal interactions between paired beta strands, 
the likelihood-maximization for a Markov random field does not
 exhibit optimal substructure, and dynamic
 programming cannot solve it quickly \cite{Menke:2010ti,Daniels:2012}.
The~purpose of MRFy is to explore new methods of \emph{approximating}
maximum likelihood.

@@

MRFy treats the beta 
strands, which are build into the model,
 as ``beads'' which can be slid along the query sequence to determine
 a \emph{placement} of the beta strands.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of any subsequence that appears
\emph{between} two beta strands can be computed exactly and
efficiently using Viterbi's algorithm.
(The result of the exact computation is a probability that is
\emph{conditioned} on the placement.)
The likelihood is computed by a function \texttt{score},
which given a model and a {placement} returns the {placement}
paired with a \emph{score}.
A~score is a negated log probability, so lower scores are better.

MRFy searches for placements stochastically,
trying to minimize scores.
We~are exploring several stochastic algorithms: random-mutation hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
\remark{It would be pleasant to connect these to the functions below,
but we probably lack room.}
Our~exploration relies on higher-order functions.

MRFy's search works with a sequence of \emph{generations}.
A~generation is a collection of scored \emph{placements}.
We~pass the
search function a \texttt{Scorer}, which is the result of partially
applying \texttt{score} to the model.
\smallverbatiminput{scoredecl}
A~search algorithm is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may not be any better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation as an additional parameter.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the list of accepted generations and the age of the current
generation. 
\end{itemize}
This interface provides ample leeway for experiments.
For~example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The interface \texttt{quit} makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.

>>>>

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality.
In~MRFy, they are used to implement new functionality.

A hidden Markov model uses local information about
amino-acid sequences.
But when a real protein folds in three dimensions, 
groups of amino acids 
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
We~are studying adjacent groups called \emph{beta strands}, which
can become hydrogen-bonded to each other,
and are effectively ``stuck together.''

A~mutation in a paired beta strand may break the pairing, in which
 case the mutated protein will not fold properly and is unlikely to
 survive natural selection.
But a mutation in a paired beta strand \emph{can} survive if it is
 accompanied by a \emph{compensatory} mutation in the strand with which it is
 paired. 
The probability of finding amino acid~$x_i$ in column~$j$ of an alignment may
 therefore depend on the amino acid in the paired column
 $\pairedwith j$.
This nonlocal interaction changes
Viterbi's equations:
each term $V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
The distance between columns~$j$ and $\pairedwith j$ can be as small
 as a few columns or as large as a few hundreds of columns.

The approach currently taken in computational biology
is~to model each pair of beta strands not with two hidden Markov
models, 
but with a pair of sequences of match states.
Each of the subsequences \emph{between} paired beta strands is then
modeled with an ordinary hidden Markov model.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}. 
Because of nonlocal interactions between paired beta strands, 
the likelihood-maximization problem for a Markov random field does not
 exhibit optimal substructure, and dynamic
 programming cannot solve it quickly \cite{Menke:2010ti,Daniels:2012}.
MRFy explores new methods of \emph{approximating}
maximum likelihood.

@@

MRFy treats the beta 
strands, which are built into the model,
 as ``beads'' which can be slid along the query sequence.
A~set of positions of beta strands is called
 a \emph{placement}.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of any subsequence that appears
\emph{between} two beta strands can be computed exactly and
efficiently using Viterbi's algorithm.
(The result of the exact computation is a probability that is
\emph{conditioned} on the placement.)

MRFy searches for placements stochastically,
trying to minimize scores.
We~are exploring several stochastic algorithms: random-mutation hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
Our~exploration relies on higher-order functions.

MRFy's search computes a sequence of \emph{generations}.
A~generation is a collection of scored {placements}.
We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the list of accepted generations and the age of the current
generation. 
\end{itemize}
This interface provides ample leeway for experiments.
For~example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The \texttt{quit} interface makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.
